```{r}
library(tidyverse)
library(BiocManager)
library(GEOquery)
library(limma)
library(CONOR)
library(janitor)
```

# 1. Recent onset T1D

## a) Extraction
```{r}

## Extraction of expressionSet and gene annotations ----
eset <- getGEO(GEO = "GSE35725")
eset <- eset[["GSE35725_series_matrix.txt.gz"]]
# exprs(eset) = log2(exprs(eset))

exprs(eset) <- backgroundCorrect(exprs(eset), method="normexp") #background correct
exprs(eset) <- normalizeBetweenArrays(object = exprs(eset)) #normalize expression profiles

anno <- as.data.frame(eset@featureData@data) #extract annotation data

expr <- as.data.frame(eset@assayData[["exprs"]]) #extract expression set

expr <- rownames_to_column(expr, var = 'ID') #copy rownames to their own column

anno_expr <- full_join(anno,expr, by = 'ID') #merge expr and anno
  
```

## b) Cleaning
```{r}

## EXTRACT/CLEAN ANNO DATA ##
anno_expr <- 
  anno_expr %>%
  select(-GB_ACC, -SPOT_ID, -`Species Scientific Name`, -`Annotation Date`, -`Sequence Type`, -`Sequence Source`, -`Target Description`, -`Representative Public ID`,-`Gene Title`, -ENTREZ_GENE_ID, -`RefSeq Transcript ID`, -`Gene Ontology Biological Process`, -`Gene Ontology Cellular Component`, -`Gene Ontology Molecular Function`) #Deselect unnecessary columns

## EXTRACT/CLEAN PHENO DATA ##
pheno_eset <-
  pData(eset) %>% 
  dplyr::select(title) %>% 
  rownames_to_column("geo_accession") %>% 
  mutate(
    title = str_remove(title, "UPN727 cells,"),
    group = str_extract(title, "autologous plasma|control|recent onset|longstanding|longitudinal")
    ) %>% 
    mutate(group = if_else(group == "recent onset", "recent_onset", 
                   if_else(group == "autologous plasma", "autologous_plasma", 
                   group)))

```

## c) Plot
```{r}
## Set-up comparisons
my_comparisons <- 
  list(
    c("control", "recent_onset"),
    c("recent_onset", "longstanding"),
    c("control", "longstanding")
  )

## Plot
anno_expr %>% 
  dplyr::rename(symbol = `Gene Symbol`) %>% 
  filter(symbol %in% c("MSTN","GDF8")) %>% 
  dplyr::select(-ID, -symbol) %>% 
  t() %>% 
  as.data.frame %>% 
  rownames_to_column("geo_accession") %>% 
  dplyr::rename(MSTN = V1) %>% 
  left_join(pheno_eset, by = "geo_accession") %>% 
  filter(!group %in% c("longitudinal", "autologous_plasma")) %>% 
  mutate(group = factor(group, levels = c("control", "recent_onset", "longstanding"))) %>% 
  ggplot(aes(x = group, y = MSTN))+
  geom_violin()+
  ggbeeswarm::geom_quasirandom()+
  ggpubr::stat_compare_means(comparisons = my_comparisons, method = "t.test", p.adjust.method = "bh")
  


```
## d) Designing model matrix and fitting linear model
```{r}
library(limma)

subset <- pheno_eset %>% select(geo_accession, group) %>% filter(!group %in% c("autologous_plasma", "longitudinal"))

subset_choices <- subset$geo_accession
vec <- subset$group

big <- anno_expr %>% dplyr::select(one_of(subset_choices),-ID, -`Gene Symbol`)

f <- factor(as.character(vec), levels = c("control","recent_onset", "longstanding"))
design <- model.matrix(~0+f)
colnames(design) <- levels(f)



fit <- lmFit(big,design)
con <- makeContrasts(
  control-recent_onset, 
  control-longstanding, 
  recent_onset-longstanding, 
  levels = design
  )
fit <- contrasts.fit(fit, contrasts=con)
fit <- eBayes(fit)

limma::plotMA(fit, coef = 1) 
limma::plotMA(fit, coef = 2) 
limma::plotMA(fit, coef = 3) 


```


## e) topTables
```{r}
library(limma)

# View(topTable(fit,sort="logFC",n=Inf, coef = 1))

gene_names <- anno %>% select(ID = ENTREZ_GENE_ID, gene_symbol = `Gene Symbol`)

## TopTable ##

tt_conVSrecent <-
  topTable(fit,sort="p",n=Inf, coef = 1, resort.by = "p") %>% 
  rownames_to_column(var = "ID") %>% 
  left_join(gene_names, by = "ID") %>% 
  select(ID, gene_symbol, everything()) %>% 
  filter(gene_symbol != "") %>% 
  mutate(FC = 2^logFC) %>% 
  mutate(FC = ifelse(logFC<0,FC*(-1),FC)) %>% 
  mutate(lab_names = if_else(gene_symbol == "MSTN", "MSTN", ""))

tt_conVSlong <-
  topTable(fit,sort="p",n=Inf, coef = 2, resort.by = "p") %>% 
  rownames_to_column(var = "ID") %>% 
  left_join(gene_names, by = "ID") %>% 
  select(ID, gene_symbol, everything()) %>% 
  filter(gene_symbol != "") %>% 
  mutate(FC = 2^logFC) %>% 
  mutate(FC = ifelse(logFC<0,FC*(-1),FC) )%>% 
  mutate(lab_names = if_else(gene_symbol == "MSTN", "MSTN", ""))

tt_recentVSlong <-
  topTable(fit,sort="p",n=Inf, coef = 3, resort.by = "p") %>% 
  rownames_to_column(var = "ID") %>% 
  left_join(gene_names, by = "ID") %>% 
  select(ID, gene_symbol, everything()) %>% 
  filter(gene_symbol != "") %>% 
  mutate(FC = 2^logFC) %>% 
  mutate(FC = ifelse(logFC<0,FC*(-1),FC)) %>% 
  mutate(lab_names = if_else(gene_symbol == "MSTN", "MSTN", ""))

  

tt_conVSrecent %>% filter(gene_symbol == "MSTN")
tt_conVSlong %>% filter(gene_symbol == "MSTN")
tt_recentVSlong %>% filter(gene_symbol == "MSTN")


```

## f) VolcanoPlot
```{r fig.height=6, fig.width=6}
library(EnhancedVolcano)

EnhancedVolcano(tt_conVSrecent,
    lab = tt_conVSrecent$gene_symbol,
    x = 'logFC',
    y = 'P.Value',
    # xlim = c(-5, 5),
    # ylim = c(0, 8),
    title = 'Recent Onset T1D',
    pCutoff = 10e-5,
    FCcutoff = 1,
    transcriptPointSize = 1.5,
    transcriptLabSize = 4.0,
    subtitle = "",
    selectLab = tt_conVSrecent$lab_names)

EnhancedVolcano(tt_conVSlong,
    lab = tt_conVSlong$gene_symbol,
    x = 'logFC',
    y = 'P.Value',
    # xlim = c(-5, 5),
    # ylim = c(0, 8),
    title = 'Longstanding T1D',
    pCutoff = 10e-5,
    FCcutoff = 1,
    transcriptPointSize = 1.5,
    transcriptLabSize = 4.0,
    subtitle = "",
    selectLab = tt_conVSlong$lab_names)


```
# 2. T2D Myotubes

## a) Extraction
```{r}

## Extraction of expressionSet and gene annotations ----
eset <- getGEO(GEO = "GSE12643")
eset <- eset[["GSE12643_series_matrix.txt.gz"]]
# exprs(eset) = log2(exprs(eset))

exprs(eset) <- backgroundCorrect(exprs(eset), method="normexp") #background correct
exprs(eset) <- normalizeBetweenArrays(object = exprs(eset)) #normalize expression profiles

anno <- as.data.frame(eset@featureData@data) #extract annotation data

expr <- as.data.frame(eset@assayData[["exprs"]]) #extract expression set

expr <- rownames_to_column(expr, var = 'ID') #copy rownames to their own column

anno_expr <- dplyr::full_join(anno,expr, by = 'ID') #merge expr and anno
  
```

## b) Cleaning
```{r}

## EXTRACT/CLEAN ANNO DATA ##
anno_expr <- 
  anno_expr %>%
  select(-GB_ACC, -SPOT_ID, -`Species Scientific Name`, -`Annotation Date`, -`Sequence Type`, -`Sequence Source`, -`Target Description`, -`Representative Public ID`,-`Gene Title`, -ENTREZ_GENE_ID, -`RefSeq Transcript ID`, -`Gene Ontology Biological Process`, -`Gene Ontology Cellular Component`, -`Gene Ontology Molecular Function`) #Deselect unnecessary columns

## EXTRACT/CLEAN PHENO DATA ##
pheno_eset <-
  pData(eset) %>% 
  dplyr::select(title) %>% 
  rownames_to_column("geo_accession") %>% 
  mutate(
    group = str_extract(title, "type 2 diabetes|Cont")) %>% 
  mutate(
    group = if_else(group == "Cont", "CON",
            if_else(group == "type 2 diabetes", "T2D",
                    group))
    )

pheno_eset
```

## c) Plot
```{r}
## Set-up comparisons
my_comparisons <- 
  list(
    c("CON", "T2D")
  )

## Plot
anno_expr %>% 
  dplyr::rename(symbol = `Gene Symbol`) %>% 
  filter(symbol %in% c("MSTN","GDF8")) %>% 
  dplyr::select(-ID, -symbol) %>% 
  t() %>% 
  as.data.frame %>% 
  rownames_to_column("geo_accession") %>% 
  dplyr::rename(MSTN = V1) %>% 
  left_join(pheno_eset, by = "geo_accession") %>% 
  ggplot(aes(x = group, y = MSTN))+
  geom_violin()+
  ggbeeswarm::geom_quasirandom()+
  ggpubr::stat_compare_means(comparisons = my_comparisons, method = "t.test", p.adjust.method = "bh")
  


```

## d) Designing model matrix and fitting linear model
```{r}
library(limma)

subset <- pheno_eset %>% select(geo_accession, group) 

subset_choices <- subset$geo_accession
vec <- subset$group

big <- anno_expr %>% dplyr::select(one_of(subset_choices),-ID, -`Gene Symbol`)

f <- factor(as.character(vec), levels = c("CON","T2D"))
design <- model.matrix(~0+f)
colnames(design) <- levels(f)



fit <- lmFit(big,design)
con <- makeContrasts(
  T2D-CON,
  levels = design
  )
fit <- contrasts.fit(fit, contrasts=con)
fit <- eBayes(fit)

limma::plotMA(fit, coef = 1) 


```


## e) topTables
```{r}
library(limma)

# View(topTable(fit,sort="logFC",n=Inf, coef = 1))

gene_names <- anno %>% select(ID = ENTREZ_GENE_ID, gene_symbol = `Gene Symbol`)

## TopTable ##

tt_conVSt2d <-
  topTable(fit,sort="p",n=Inf, coef = 1, resort.by = "p") %>% 
  rownames_to_column(var = "ID") %>% 
  left_join(gene_names, by = "ID") %>% 
  select(ID, gene_symbol, everything()) %>% 
  filter(gene_symbol != "") %>% 
  mutate(FC = 2^logFC) %>% 
  mutate(FC = ifelse(logFC<0,FC*(-1),FC)) %>% 
  mutate(lab_names = if_else(gene_symbol == "MSTN", "MSTN", ""))



  

tt_conVSt2d %>% filter(gene_symbol == "MSTN")


```

## f) VolcanoPlot
```{r fig.height=6, fig.width=6}
library(EnhancedVolcano)

EnhancedVolcano(tt_conVSt2d,
    lab = tt_conVSt2d$gene_symbol,
    x = 'logFC',
    y = 'P.Value',
    # xlim = c(-5, 5),
    # ylim = c(0, 8),
    title = 'T2D vs CON',
    pCutoff = 10e-5,
    FCcutoff = 1,
    transcriptPointSize = 1.5,
    transcriptLabSize = 4.0,
    subtitle = "",
    selectLab = tt_conVSt2d$lab_names)

```
# 3. Myotubes after HFD

## a) Extraction
```{r}

## Extraction of expressionSet and gene annotations ----
eset <- getGEO(GEO = "GSE68231")
eset <- eset[["GSE68231_series_matrix.txt.gz"]]
# exprs(eset) = log2(exprs(eset))

exprs(eset) <- backgroundCorrect(exprs(eset), method="normexp") #background correct
exprs(eset) <- normalizeBetweenArrays(object = exprs(eset)) #normalize expression profiles

anno <- as.data.frame(eset@featureData@data) #extract annotation data

expr <- as.data.frame(eset@assayData[["exprs"]]) #extract expression set

expr <- rownames_to_column(expr, var = 'ID') #copy rownames to their own column

anno_expr <- dplyr::full_join(anno,expr, by = 'ID') #merge expr and anno
  
```

## b) Cleaning
```{r}

## EXTRACT/CLEAN ANNO DATA ##
anno_expr <- 
  anno_expr %>%
  select(-GB_ACC, -SPOT_ID, -`Species Scientific Name`, -`Annotation Date`, -`Sequence Type`, -`Sequence Source`, -`Target Description`, -`Representative Public ID`,-`Gene Title`, -ENTREZ_GENE_ID, -`RefSeq Transcript ID`, -`Gene Ontology Biological Process`, -`Gene Ontology Cellular Component`, -`Gene Ontology Molecular Function`) #Deselect unnecessary columns

## EXTRACT/CLEAN PHENO DATA ##
pheno_eset <-
  pData(eset) %>% 
  dplyr::select(title) %>% 
  rownames_to_column("geo_accession") %>% 
  mutate(group = str_extract(title, "group_...")) %>% 
  mutate(group = str_remove(group,".$")) %>% 
  rownames_to_column("rowname") %>% 
  mutate(vec = paste(rowname, group, sep = "_"))


pheno_eset
```

## c) Plot
```{r}
## Set-up comparisons
my_comparisons <- 
  list(
    c("group_HR", "group_LR")
  )

## Plot
anno_expr %>% 
  dplyr::rename(symbol = `Gene Symbol`) %>% 
  filter(symbol %in% c("MSTN","GDF8")) %>% 
  dplyr::select(-ID, -symbol) %>% 
  t() %>% 
  as.data.frame %>% 
  rownames_to_column("geo_accession") %>% 
  dplyr::rename(MSTN = V1) %>% 
  left_join(pheno_eset, by = "geo_accession") %>% 
  ggplot(aes(x = group, y = MSTN))+
  geom_violin()+
  ggbeeswarm::geom_quasirandom()+
  ggpubr::stat_compare_means(comparisons = my_comparisons, method = "t.test", p.adjust.method = "bh")
  


```

## d) Designing model matrix and fitting linear model
```{r}
library(limma)

vec <- pheno_eset %>% select(vec) %>% mutate(vec = word(vec, start = 2, end = 3, sep = "_")) %>% pull

big <- anno_expr %>% dplyr::select(-ID, -`Gene Symbol`)

f <- factor(as.character(vec), levels = c("group_LR","group_HR"))
design <- model.matrix(~0+f)
colnames(design) <- c("group_LR","group_HR")

fit <- lmFit(big,design)
con <- makeContrasts(group_HR-group_LR, levels = design)
fit <- contrasts.fit(fit, contrasts=con)
fit <- eBayes(fit)

limma::plotMA(fit, coef = 1) 


```


## e) topTables
```{r}
library(limma)

# View(topTable(fit,sort="logFC",n=Inf, coef = 1))

gene_names <- anno %>% select(ID = ENTREZ_GENE_ID, gene_symbol = `Gene Symbol`)

## TopTable ##

tt <-
  topTable(fit,sort="p",n=Inf, coef = 1, resort.by = "p") %>% 
  rownames_to_column(var = "ID") %>% 
  left_join(gene_names, by = "ID") %>% 
  select(ID, gene_symbol, everything()) %>% 
  filter(gene_symbol != "") %>% 
  mutate(FC = 2^logFC) %>% 
  mutate(FC = ifelse(logFC<0,FC*(-1),FC))
  




```

## f) VolcanoPlot
```{r fig.height=6, fig.width=6}
library(EnhancedVolcano)

EnhancedVolcano(tt,
    lab = tt$gene_symbol,
    x = 'logFC',
    y = 'P.Value',
    # xlim = c(-5, 5),
    # ylim = c(0, 8),
    title = 'HFD',
    pCutoff = 10e-5,
    FCcutoff = 1,
    transcriptPointSize = 1.5,
    transcriptLabSize = 4.0,
    subtitle = "")


```

# 4. Type 2 diabetes and insulin resistance

## a) Extraction
```{r}

## Extraction of expressionSet and gene annotations ----
eset <- getGEO(GEO = "GSE22309")
eset <- eset[["GSE22309_series_matrix.txt.gz"]]
# exprs(eset) = log2(exprs(eset))

exprs(eset) <- backgroundCorrect(exprs(eset), method="normexp") #background correct
exprs(eset) <- normalizeBetweenArrays(object = exprs(eset)) #normalize expression profiles

anno <- as.data.frame(eset@featureData@data) #extract annotation data

expr <- as.data.frame(eset@assayData[["exprs"]]) #extract expression set

expr <- rownames_to_column(expr, var = 'ID') #copy rownames to their own column

anno_expr <- dplyr::full_join(anno,expr, by = 'ID') #merge expr and anno
  
```

## b) Cleaning
```{r}

## EXTRACT/CLEAN ANNO DATA ##
anno_expr <- 
  anno_expr %>%
  select(-GB_ACC, -SPOT_ID, -`Species Scientific Name`, -`Annotation Date`, -`Sequence Type`, -`Sequence Source`, -`Target Description`, -`Representative Public ID`,-`Gene Title`, -ENTREZ_GENE_ID, -`RefSeq Transcript ID`, -`Gene Ontology Biological Process`, -`Gene Ontology Cellular Component`, -`Gene Ontology Molecular Function`) #Deselect unnecessary columns

## EXTRACT/CLEAN PHENO DATA ##
pheno_eset <-
  pData(eset) %>% 
  dplyr::select(title) %>% 
  rownames_to_column("geo_accession") %>% 
  mutate(group = str_extract(title, "IS.*|IR.*|DB.*")) %>% 
  mutate(vec = str_replace(group," ", "_"))

pheno_eset
```

## c) Plot
```{r}
## Set-up comparisons

groups <- pheno_eset %>% distinct(vec) %>% pull

my_comparisons <- 
  list(
    c("IS_basal", "IS_stimulate"),
    c("IR_basal", "IR_stimulate"),
    c("DB_basal", "DB_stimulate")
  )

my_comparisons_basal <- 
  list(
    c("IS_basal", "IR_basal"),
    c("DB_basal", "IS_basal"),
    c("DB_basal", "IR_basal")
  )

my_comparisons_stim <- 
  list(
    c("IS_stimulate", "IR_stimulate"),
    c("DB_stimulate", "IS_stimulate"),
    c("DB_stimulate", "IR_stimulate")
  )


## Plot
anno_expr %>% 
  dplyr::rename(symbol = `Gene Symbol`) %>% 
  filter(symbol %in% c("MSTN","GDF8")) %>% 
  dplyr::select(-ID, -symbol) %>% 
  t() %>% 
  as.data.frame %>% 
  rownames_to_column("geo_accession") %>% 
  dplyr::rename(MSTN = V1) %>% 
  left_join(pheno_eset, by = "geo_accession") %>% 
  ggplot(aes(x = vec, y = MSTN))+
  geom_violin()+
  ggbeeswarm::geom_quasirandom()+
  ggpubr::stat_compare_means(comparisons = my_comparisons_basal, method = "t.test", p.adjust.method = "bh")
  


```

## d) Designing model matrix and fitting linear model
```{r}
library(limma)

vec <- pheno_eset$vec
big <- anno_expr %>% dplyr::select(-ID, -`Gene Symbol`)

f <- factor(as.character(vec), levels = groups)
design <- model.matrix(~0+f)
colnames(design) <- groups

fit <- lmFit(big,design)

con <- 
  makeContrasts(
    DB_basal-IS_basal, # coef 1
    DB_basal-IR_basal, # coef 2
    IR_basal-IS_basal, # coef 3
    DB_stimulate-IS_stimulate, # coef 4
    DB_stimulate-IR_stimulate, # coef 5
    IR_stimulate-IS_stimulate, # coef 6
    levels = design
  )

fit <- contrasts.fit(fit, contrasts=con)
fit <- eBayes(fit)

limma::plotMA(fit, coef = 1) 
limma::plotMA(fit, coef = 2) 
limma::plotMA(fit, coef = 3) 
limma::plotMA(fit, coef = 4) 
limma::plotMA(fit, coef = 5) 
limma::plotMA(fit, coef = 6) 

groups

```


## e) topTables
```{r}
library(limma)

# View(topTable(fit,sort="logFC",n=Inf, coef = 1))

gene_names <- anno %>% dplyr::select(ID = ENTREZ_GENE_ID, gene_symbol = `Gene Symbol`)

## TopTable ##


tt <- function(coef){
  topTable(fit,sort="p",n=Inf, coef = coef, resort.by = "p") %>% 
  rownames_to_column(var = "ID") %>% 
  left_join(gene_names, by = "ID") %>% 
  select(ID, gene_symbol, everything()) %>% 
  filter(gene_symbol != "") %>% 
  mutate(
    FC = 2^logFC,
    FC = ifelse(logFC<0,FC*(-1),FC),
    lab_values = if_else(gene_symbol == "MSTN|GDF8", "MSTN", "")
    )
}

# DB_basal-IS_basal, # coef 1
tt1 <- tt(1) 
# DB_basal-IR_basal, # coef 2
tt2 <- tt(2)
# IR_basal-IS_basal, # coef 3
tt3 <- tt(3)
# DB_stimulate-IS_stimulate, # coef 4
tt4 <- tt(4)
# DB_stimulate-IR_stimulate, # coef 5
tt5 <- tt(5)
# IR_stimulate-IS_stimulate, # coef 6
tt6 <- tt(6)



```

## f) VolcanoPlot
```{r fig.height=6, fig.width=6}
library(EnhancedVolcano)

volc <- function(tt, title){
  EnhancedVolcano(tt,
    lab = tt$gene_symbol,
    x = 'logFC',
    y = 'P.Value',
    title = title,
    pCutoff = 10e-6,
    FCcutoff = 1,
    transcriptPointSize = 1.5,
    transcriptLabSize = 4.0,
    subtitle = "")
}

volc(tt1, "DB basal vs IS basal")
volc(tt2, "DB basal vs IR basal")
volc(tt3, "IR basal vs IS basal")
volc(tt4, "DB stimulate vs IS stimulate")
volc(tt5, "DB stimulate vs IR stimulate")
volc(tt6, "IR stimulate vs IS stimulate")

```

# 5. Development of T1D 

## a) Extraction
```{r}

## Extraction of expressionSet and gene annotations ----
eset <- getGEO(GEO = "GSE30210")
eset <- eset[["GSE30210_series_matrix.txt.gz"]]

exprs(eset) <- backgroundCorrect(exprs(eset), method="normexp") #background correct
exprs(eset) <- normalizeBetweenArrays(object = exprs(eset)) #normalize expression profiles

anno <- as.data.frame(eset@featureData@data) #extract annotation data

expr <- as.data.frame(eset@assayData[["exprs"]]) #extract expression set

expr <- rownames_to_column(expr, var = 'ID') #copy rownames to their own column

anno_expr <- dplyr::full_join(anno,expr, by = 'ID') %>% #merge expr and anno
  select(ID, ILMN_Gene) # select id columns
  
```

## b) Cleaning
```{r}

## EXTRACT/CLEAN PHENO DATA ##
pheno_eset <-
  pData(eset) %>% 
  dplyr::select(title) %>% 
  rownames_to_column("geo_accession") %>% 
  mutate(
    group = str_extract(title, "Case|Control"),
    pair_no = str_extract(title, "[:digit:]{1}(?=_)"),
    timepoint = word(title, 2, sep = "_")
    )


pheno_eset
```

## c) Plot
```{r fig.width=15, fig.height=5}
## Set-up comparisons

groups <- pheno_eset %>% distinct(vec) %>% pull

my_comparisons <- 
  list(
    c("1", "2"),   
    c("2", "3"),  
    c("3", "4"),  
    c("4", "5"),    
    c("5", "6"),    
    c("6", "7")    
  )

my_comparisons_group <- 
  list(
    c("Control", "Case")
  )



## Plot
anno_expr %>% 
  dplyr::rename(symbol = ILMN_Gene) %>% 
  filter(symbol %in% c("MSTN","GDF8")) %>% 
  dplyr::select(GSM747675:GSM747921) %>% 
  t() %>% 
  as.data.frame %>% 
  rownames_to_column("geo_accession") %>% 
  dplyr::rename(MSTN = V1) %>% 
  left_join(pheno_eset, by = "geo_accession") %>% 
  filter(timepoint %in% c("1", "2", "3", "4", "5", "6", "7")) %>% 
  mutate(group = factor(group, levels = c("Control", "Case"))) %>% 
  
  ggplot(aes(x = group, y = MSTN, color = group))+
  geom_violin()+
  ggbeeswarm::geom_quasirandom(dodge.width = 0.8)+
  ggpubr::stat_compare_means(comparisons = my_comparisons_group, method = "t.test", p.adjust.method = "bh")
  


```

# 6. T1D vs T2D

## a) Extraction
```{r}

## Extraction of expressionSet and gene annotations ----
eset <- getGEO(GEO = "GSE50866")
eset <- eset[["GSE50866_series_matrix.txt.gz"]]
# exprs(eset) = log2(exprs(eset))

exprs(eset) <- backgroundCorrect(exprs(eset), method="normexp") #background correct
exprs(eset) <- normalizeBetweenArrays(object = exprs(eset)) #normalize expression profiles

anno <- as.data.frame(eset@featureData@data) #extract annotation data

expr <- 
  as.data.frame(eset@assayData[["exprs"]]) %>% #extract expression set
  rownames_to_column(var = 'ID') %>%  #copy rownames to their own column
  mutate(ID = as.numeric(ID))

anno_expr <-
  dplyr::full_join(anno,expr, by = 'ID') %>%  #merge expr and anno
  select(ID, Description, GSM1231310:GSM1231368) %>% 
  mutate(symbol = word(Description, -1, sep = " ")) %>% 
  mutate(symbol = str_replace(symbol, "\\(", "")) %>% 
  mutate(symbol = str_replace(symbol, "\\)$", "")) %>% 
  select(symbol, everything())

```

## b) Cleaning
```{r}

## EXTRACT/CLEAN PHENO DATA ##
pheno_eset <-
  pData(eset) %>% 
  dplyr::select(title) %>% 
  rownames_to_column("geo_accession") %>% 
  mutate(group = str_extract(title, "NGT|T1DM|T2DM"),
         vec = group)

pheno_eset
```

## c) Plot
```{r}
## Set-up comparisons
my_comparisons <- 
  list(
    c("NGT", "T1DM"),
    c("NGT", "T2DM"),
    c("T1DM", "T2DM")
  )

## Plot
anno_expr %>% 
  filter(symbol %in% c("MSTN","GDF8")) %>% 
  dplyr::select(-ID, -symbol, -Description) %>% 
  t() %>% 
  as.data.frame %>% 
  rownames_to_column("geo_accession") %>% 
  dplyr::rename(MSTN = V1) %>% 
  left_join(pheno_eset, by = "geo_accession") %>% 
  mutate(group = factor(group, levels = c("NGT", "T1DM", "T2DM"))) %>% 
  
  ggplot(aes(x = group, y = MSTN))+
  geom_violin()+
  ggbeeswarm::geom_quasirandom()+
  ggpubr::stat_compare_means(comparisons = my_comparisons, method = "t.test", p.adjust.method = "bh")
  


```

## d) Designing model matrix and fitting linear model
```{r}
library(limma)

vec <- pheno_eset %>% select(vec) %>% mutate(vec = word(vec, start = 2, end = 3, sep = "_")) %>% pull

big <- anno_expr %>% dplyr::select(-ID, -`Gene Symbol`)

f <- factor(as.character(vec), levels = c("group_LR","group_HR"))
design <- model.matrix(~0+f)
colnames(design) <- c("group_LR","group_HR")

fit <- lmFit(big,design)
con <- makeContrasts(group_HR-group_LR, levels = design)
fit <- contrasts.fit(fit, contrasts=con)
fit <- eBayes(fit)

limma::plotMA(fit, coef = 1) 


```


## e) topTables
```{r}
library(limma)

# View(topTable(fit,sort="logFC",n=Inf, coef = 1))

gene_names <- anno %>% select(ID = ENTREZ_GENE_ID, gene_symbol = `Gene Symbol`)

## TopTable ##

tt <-
  topTable(fit,sort="p",n=Inf, coef = 1, resort.by = "p") %>% 
  rownames_to_column(var = "ID") %>% 
  left_join(gene_names, by = "ID") %>% 
  select(ID, gene_symbol, everything()) %>% 
  filter(gene_symbol != "") %>% 
  mutate(FC = 2^logFC) %>% 
  mutate(FC = ifelse(logFC<0,FC*(-1),FC))
  




```

## f) VolcanoPlot
```{r fig.height=6, fig.width=6}
library(EnhancedVolcano)

EnhancedVolcano(tt,
    lab = tt$gene_symbol,
    x = 'logFC',
    y = 'P.Value',
    # xlim = c(-5, 5),
    # ylim = c(0, 8),
    title = 'HFD',
    pCutoff = 10e-5,
    FCcutoff = 1,
    transcriptPointSize = 1.5,
    transcriptLabSize = 4.0,
    subtitle = "")


```
