```{r}
library(tidyverse)
library(BiocManager)
library(GEOquery)
library(affy)
library(limma)
library(CONOR)
library(janitor)
library(patchwork)
options(scipen = 999)

## HELPER FUNCTIONS ----

```


# 1. GSE22039 - Wu

## a) Extraction
```{r}
###########
## SETUP ##
###########

gse <- "GSE22309"

## Extraction of expressionSet and QC ----
eset <- getGEO(GEO = paste0(gse))
eset <- eset[[paste0(gse,"_series_matrix.txt.gz")]]

expr <- exprs(eset)
expr <- expr %>% as.data.frame %>% rownames_to_column(var = 'ID') # copy rownames to their own column

## QC ##
# Boxplots
expr %>%
  gather(gsm, value, -ID) %>% 
  ggplot(aes(x = gsm, y = value))+
  geom_boxplot()
# Histogram
expr %>%
  gather(gsm, value, -ID) %>% 
  ggplot(aes(x =value))+
  geom_histogram()


## Normalize if necessary ---
# expr <- exprs(eset)
# expr <- backgroundCorrect(expr, method= "normexp") # background correct
# expr <- normalizeBetweenArrays(object = expr) # normalize expression profiles
# expr <- log2(expr) %>% as.data.frame 
# expr <- expr %>% as.data.frame %>% rownames_to_column(var = 'ID') # copy rownames to their own column

## Annotate data expressionSet
anno <- 
  as.data.frame(eset@featureData@data) %>%  # extract annotation data
    dplyr::select(ID, `Gene Symbol`, ENTREZ_GENE_ID)

anno_expr <- dplyr::full_join(anno,expr, by = 'ID') # merge expr and anno
   
```

## b) Cleaning
```{r}

## EXTRACT/CLEAN PHENO DATA ##
pheno_eset <-
  pData(eset) %>% 
  dplyr::select(title, `choa:ch1`:`perfat:ch1`) %>% 
  rownames_to_column("geo_accession") %>% 
  mutate(group = str_extract(title, "IS.*|IR.*|DB.*")) %>% 
  mutate(vec = str_replace(group," ", "_"))

pheno_eset
```

## c) Plot
```{r, fig.width = 10, fig.height=10}
## Set-up comparisons

groups <- pheno_eset %>% distinct(vec) %>% pull

my_comparisons <- 
  list(
    c("CON", "IR"),
    c("CON", "T2D")
  )



## Plot data

meta_vars <- pheno_eset %>% dplyr::select(`choa:ch1`:`perfat:ch1`) %>% names

plot_data <-
  anno_expr %>% 
  dplyr::rename(symbol = `Gene Symbol`) %>% 
  filter(symbol %in% c("MSTN","GDF8")) %>% 
  dplyr::select(-ID, -symbol, -ENTREZ_GENE_ID) %>% 
  t() %>% 
  as.data.frame %>% 
  rownames_to_column("geo_accession") %>% 
  dplyr::rename(MSTN = V1) %>% 
  left_join(pheno_eset, by = "geo_accession") %>% 
   mutate(group = str_replace(group, "IS basal|IS stimulate", "CON")) %>%
   mutate(group = str_replace(group, "IR basal|IR stimulate", "IR")) %>%
   mutate(group = str_replace(group, "DB basal|DB stimulate", "T2D")) %>% 
  mutate_at(.vars = meta_vars, .funs = as.numeric) %>% 
  group_by(group) %>% 
  fill(meta_vars) %>% 
  ungroup

mean_value <-
  plot_data %>% 
    filter(group == "CON") %>% 
    summarise(MSTN = mean(MSTN, na.rm = T)) %>% 
    pull

## PLOT ##

p1 <- 
  plot_data %>% 
  filter(str_detect(vec, "_basal")) %>% 
  ggplot(aes(x = group, y = MSTN+(2*abs(mean_value))))+
   stat_summary(fun.data = mean_se, geom = "errorbar", width=0.2, size = 1, color = "#000000") +
   stat_summary(fun.y = mean, geom = "bar", width=0.8, size = 1, color = "#000000", fill=c("#ffffff", "grey60", "black")) +
   theme_classic() +
   scale_y_continuous(expand = c(0, 0)) +
   scale_color_manual(values=c("#000000","#000000")) +
   ggbeeswarm::geom_quasirandom(size = 3, stroke = 1, width = 0.2, shape = 21, fill = "white")+
    theme(aspect.ratio = 1,
          plot.title = element_text(vjust= 7.5),
          axis.title.x = element_blank(),
          axis.text.x = element_text(color = "black"),
          axis.title.y = element_text(vjust = 4.5),
          axis.text.y = element_text(color = "black"),
          legend.position = "none",
          axis.ticks.x = element_blank(),
          plot.margin = unit(c(20,20,20,20),"pt")) + 
   labs(y=expression(paste(italic("MSTN "), log[2], " mRNA expression")),
        x="Group") + 
    ggpubr::stat_compare_means(aes(label = ..p.adj..), comparisons = my_comparisons, method = "t.test", hide.ns = T) +
    coord_cartesian(clip = "off")


p1

```

## d) Designing model matrix and fitting linear model
```{r}
library(limma)

vec <- pheno_eset$vec
big <- anno_expr %>% dplyr::select(-ID, -`Gene Symbol`, -ENTREZ_GENE_ID)

f <- factor(as.character(vec), levels = groups)
design <- model.matrix(~0+f)
colnames(design) <- groups

fit <- lmFit(big,design)

con <- 
  makeContrasts(
   DB_basal - IS_basal, # coef 1
   IR_basal - IS_basal, # coef 2
    levels = design
  )

fit <- contrasts.fit(fit, contrasts=con)
fit <- eBayes(fit)

limma::plotMA(fit, coef = 1) 
limma::plotMA(fit, coef = 2) 


```


## e) topTables
```{r}
library(limma)

gene_names <- anno %>% dplyr::select(ID = ENTREZ_GENE_ID, gene_symbol = `Gene Symbol`)

## TopTable ##
tt <- function(coef){
  topTable(fit,sort="p",n=Inf, coef = coef, resort.by = "p") %>% 
  rownames_to_column(var = "ID") %>% 
  left_join(gene_names, by = "ID") %>% 
  dplyr::select(ID, gene_symbol, everything()) %>% 
  filter(gene_symbol != "") %>% 
  mutate(
    FC = 2^logFC,
    FC = ifelse(logFC<0,FC*(-1),FC),
    lab_values = str_extract(gene_symbol, "MSTN|GDF8")
    )
}

# DB_basal-IS_basal, # coef 1
tt1 <- tt(1) 
# IR_basal-IS_basal, # coef 2
tt2 <- tt(2)


CONvT2D <- tt1 %>% filter(gene_symbol == "MSTN")
CONvIR <- tt2 %>% filter(gene_symbol == "MSTN")

tt_Wu <- 
  list(
  CONvT2D,
  CONvIR
  ) %>%  
  setNames(c("CONvT2D", "CONvIR")) %>%
    bind_rows(.id = "comparison")

```

## f) VolcanoPlot
```{r fig.height=6, fig.width=6}
library(EnhancedVolcano)

volc <- function(tt, title){
  EnhancedVolcano(tt,
    lab = tt$gene_symbol,
    x = 'logFC',
    y = 'P.Value',
    title = title,
    pCutoff = 10e-6,
    FCcutoff = 1,
    transcriptPointSize = 1.5,
    transcriptLabSize = 4.0,
    subtitle = "", 
    selectLab = tt$lab_values)
}

volc(tt1, "DB basal vs IS basal")
volc(tt2, "DB basal vs IR basal")

```
## g) Pull relevant data
```{r}

plot_data_Wu <- 
  plot_data %>% 
  dplyr::select(geo_accession, MSTN, group)

```


# 2. GSE18732 - Gallagher

## a) Extraction
```{r}
###########
## SETUP ##
###########

gse <- "GSE18732"


## Extraction of PROCESSED expressionSet ----
eset_proc <- getGEO(GEO = paste0(gse))
eset_proc <- eset_proc[[paste0(gse,"_series_matrix.txt.gz")]]

# Reading raw data
filePaths <-  getGEOSuppFiles(paste0(gse))
untar(paste0(gse,"/",gse,"_RAW.tar"), exdir=paste(gse,"data",sep = "/"))
cels <- list.files(paste0(gse,"/data/"), pattern = "[gz]")
sapply(paste(gse,"data", cels, sep="/"), gunzip)
celpath <- paste(gse,"data/", sep = "/")
fns <- list.celfiles(path=celpath,full.names=TRUE)
fns
cat("Reading files:\n",paste(fns,collapse="\n"),"\n")
 
#Loading the CEL-files into an AffyBatch object
celfiles <- ReadAffy(celfile.path=celpath)
oligo::boxplot(celfiles)

## RMA-normalization
eset <- rma(celfiles)
oligo::boxplot(eset)

## Extract expression matrix 
expr <- exprs(eset)
expr <- expr %>% as.data.frame %>% rownames_to_column(var = 'ID') # copy rownames to their own column

## Get annotation data for platform
library(hgu133plus2.db)
anno_db <- select(hgu133plus2.db, keys = keys(hgu133plus2.db), columns=c("SYMBOL","GENENAME", "ENTREZID"), keytype="PROBEID") %>% 
  dplyr::rename(ID = PROBEID)


## Annotate data expressionSet
anno <-
  as.data.frame(eset@featureData@data) %>%
    rownames_to_column("ID") %>% 
    left_join(anno_db, by = "ID")
  
anno_expr <- dplyr::full_join(anno,expr, by = 'ID') # merge expr and anno
   

```

## b) Cleaning
```{r}

## EXTRACT/CLEAN PHENO DATA ##

pData(eset) <- pData(eset_proc)

pheno_eset <-
  pData(eset) %>% 
  dplyr::select(title) %>% 
  rownames_to_column("geo_accession") %>% 
  mutate(group = str_extract(title, "normal|glucoseIntolerant|diabetic")) %>% 
  mutate(group = recode(group, "glucoseIntolerant" = "IR", "normal" = "CON", "diabetic" = "T2D"),
         geo_accession = paste0(geo_accession,".CEL"))

pheno_eset
```

## c) Plot
```{r, fig.width = 10, fig.height=10}
## Set-up comparisons

groups <- pheno_eset %>% distinct(vec) %>% pull

my_comparisons <- 
  list(
    c("CON", "IR"),
    c("CON", "T2D")
  )



## Plot data

# meta_vars <- plot_data %>% select(`choa:ch1`:`perfat:ch1`) %>% names

plot_data <-
  anno_expr %>% 
  filter(str_detect(SYMBOL, "MSTN")) %>% 
  dplyr::select(GSM465274.CEL:GSM465391.CEL) %>% 
  t() %>% 
  as.data.frame %>% 
  rownames_to_column("geo_accession") %>% 
  dplyr::rename(MSTN = V1) %>% 
  left_join(pheno_eset, by = "geo_accession") 

## PLOT ##

plot_data %>% 
  ggplot(aes(x = group, y = MSTN))+
   stat_summary(fun.data = mean_se, geom = "errorbar", width=0.2, size = 1, color = "#000000") +
   stat_summary(fun.y = mean, geom = "bar", width=0.8, size = 1, color = "#000000", fill=c("#ffffff", "grey60", "black")) +
   theme_classic() +
   scale_y_continuous(expand = c(0, 0)) +
   scale_color_manual(values=c("#000000","#000000")) +
   ggbeeswarm::geom_quasirandom(size = 3, stroke = 1, width = 0.2, shape = 21, fill = "white")+
    theme(aspect.ratio = 1,
          plot.title = element_text(vjust= 7.5),
          axis.title.x = element_blank(),
          axis.text.x = element_text(color = "black"),
          axis.title.y = element_text(vjust = 4.5),
          axis.text.y = element_text(color = "black"),
          legend.position = "none",
          axis.ticks.x = element_blank(),
          plot.margin = unit(c(20,20,20,20),"pt")) + 
   labs(y=expression(paste(italic("MSTN "), log[2], " mRNA expression")),
        x="Group") + 
    ggpubr::stat_compare_means(aes(label = ..p.adj..), comparisons = my_comparisons, method = "wilcox.test", hide.ns = T, p.adjust.method = "fdr") +
    coord_cartesian(clip = "off")


```

## d) Designing model matrix and fitting linear model
```{r}
library(limma)

groups <- c("CON", "IR", "T2D")

vec <- pheno_eset$group
big <- anno_expr %>% dplyr::select(GSM465274.CEL:GSM465391.CEL)

f <- factor(as.character(vec), levels = groups)
design <- model.matrix(~0+f)
colnames(design) <- groups

fit <- lmFit(big,design)

con <- 
  makeContrasts(
    T2D-CON, # coef 1
    IR-CON, # coef 2
    levels = design
  )

fit <- contrasts.fit(fit, contrasts=con)
fit <- eBayes(fit)

limma::plotMA(fit, coef = 1) 
limma::plotMA(fit, coef = 2) 

```


## e) topTables
```{r}
library(limma)

gene_names <- anno %>% dplyr::select(ID = ENTREZID, gene_symbol = SYMBOL)

## TopTable ##
tt <- function(coef){
  topTable(fit,sort="p",n=Inf, coef = coef, resort.by = "p") %>% 
  rownames_to_column(var = "ID") %>% 
  left_join(gene_names, by = "ID") %>%
  dplyr::select(ID, gene_symbol, everything()) %>%
  filter(gene_symbol != "") %>%
  mutate(
    FC = 2^logFC,
    FC = ifelse(logFC<0,FC*(-1),FC),
    # lab_values = str_extract(gene_symbol, "MSTN|GDF8")
    )
}

topTable(fit,sort="p",n=Inf, coef = 1, resort.by = "p")

# CON v T2D, # coef 1
tt1 <- tt(1) 
# CON v IR, # coef 3
tt2 <- tt(2)


CONvT2D <- tt1 %>% filter(gene_symbol == "MSTN")
CONvIR <- tt2 %>% filter(gene_symbol == "MSTN")

tt_Gallagher <- 
  list(
  CONvT2D,
  CONvIR
  ) %>%  
  setNames(c("CONvT2D", "CONvIR")) %>%
    bind_rows(.id = "comparison")

```

## f) VolcanoPlot
```{r fig.height=6, fig.width=6}
library(EnhancedVolcano)

volc <- function(tt, title){
  EnhancedVolcano(tt,
    lab = tt$gene_symbol,
    x = 'logFC',
    y = 'P.Value',
    title = title,
    pCutoff = 10e-6,
    FCcutoff = 1,
    transcriptPointSize = 1.5,
    transcriptLabSize = 4.0,
    subtitle = "", 
    selectLab = tt$lab_values)
}

volc(tt1, "CON v T2D")
volc(tt2, "IR v T2D")

```

## g) Pull relevant data
```{r}

plot_data_Gallagher <- 
  plot_data %>% 
  dplyr::select(geo_accession, MSTN, group)

```

# 3. GSE19420 - vanTienen


## a) Extraction
```{r}
###########
## SETUP ##
###########

gse <- "GSE19420"

## Extraction of PROCESSED expressionSet ----
eset_proc <- getGEO(GEO = paste0(gse))
eset_proc <- eset_proc[[paste0(gse,"_series_matrix.txt.gz")]]

# Reading raw data
filePaths <-  getGEOSuppFiles(paste0(gse))
untar(paste0(gse,"/",gse,"_RAW.tar"), exdir=paste(gse,"data",sep = "/"))
cels <- list.files(paste0(gse,"/data/"), pattern = "[gz]")
sapply(paste(gse,"data", cels, sep="/"), gunzip)
celpath <- paste(gse,"data/", sep = "/")
fns <- list.celfiles(path=celpath,full.names=TRUE)
fns
cat("Reading files:\n",paste(fns,collapse="\n"),"\n")
 
#Loading the CEL-files into an AffyBatch object
celfiles <- ReadAffy(celfile.path=celpath)
oligo::boxplot(celfiles)

## RMA-normalization
eset <- rma(celfiles)
oligo::boxplot(eset)

## Extract expression matrix 
expr <- exprs(eset)
expr <- expr %>% as.data.frame %>% rownames_to_column(var = 'ID') # copy rownames to their own column

## Get annotation data for platform
anno_db <- select(hgu133plus2.db, keys = keys(hgu133plus2.db), columns=c("SYMBOL","GENENAME", "ENTREZID"), keytype="PROBEID") %>% 
  dplyr::rename(ID = PROBEID)


## Annotate data expressionSet
anno <-
  as.data.frame(eset@featureData@data) %>%
    rownames_to_column("ID") %>% 
    left_join(anno_db, by = "ID")
  
anno_expr <- dplyr::full_join(anno,expr, by = 'ID') # merge expr and anno
```

## b) Cleaning
```{r}

pData(eset) <- pData(eset_proc)

## EXTRACT/CLEAN PHENO DATA ##
pheno_eset <-
  pData(eset) %>% 
  dplyr::select(group = `group:ch1`) %>% 
  rownames_to_column("geo_accession") %>% 
  filter(group != "T2DM T52") %>% 
  mutate(group = recode(group, "T2DM T0" = "T2D", "IGT" = "CON", "NGT" = "IR"),
         geo_accession = paste0(geo_accession,".CEL")) 
  
  
pheno_eset
```

## c) Plot
```{r, fig.width = 10, fig.height=10}
## Set-up comparisons

groups <- pheno_eset %>% distinct(group) %>% pull

my_comparisons <- 
  list(
    c("CON", "IR"),
    c("CON", "T2D")
  )

## Plot data


plot_data <-
  anno_expr %>% 
  filter(str_detect(SYMBOL, "MSTN")) %>% 
  dplyr::select(GSM482933.CEL:GSM482974.CEL) %>% 
  t() %>% 
  as.data.frame %>% 
  rownames_to_column("geo_accession") %>% 
  dplyr::rename(MSTN = V1) %>% 
  left_join(pheno_eset, by = "geo_accession") %>% 
  filter(!is.na(group))

## PLOT ##

plot_data %>% 
  ggplot(aes(x = group, y = MSTN))+
   stat_summary(fun.data = mean_se, geom = "errorbar", width=0.2, size = 1, color = "#000000") +
   stat_summary(fun.y = mean, geom = "bar", width=0.8, size = 1, color = "#000000", fill=c("#ffffff", "grey60", "black")) +
   theme_classic() +
   scale_y_continuous(expand = c(0, 0)) +
   scale_color_manual(values=c("#000000","#000000")) +
   ggbeeswarm::geom_quasirandom(size = 3, stroke = 1, width = 0.2, shape = 21, fill = "white")+
    theme(aspect.ratio = 1,
          plot.title = element_text(vjust= 7.5),
          axis.title.x = element_blank(),
          axis.text.x = element_text(color = "black"),
          axis.title.y = element_text(vjust = 4.5),
          axis.text.y = element_text(color = "black"),
          legend.position = "none",
          axis.ticks.x = element_blank(),
          plot.margin = unit(c(20,20,20,20),"pt")) + 
   labs(y=expression(paste(italic("MSTN "), log[2], " mRNA expression")),
        x="Group") + 
    ggpubr::stat_compare_means(aes(label = ..p.adj..), comparisons = my_comparisons, method = "t.test", hide.ns = T, p.adjust.method = "fdr") +
    coord_cartesian(clip = "off")

```

## d) Designing model matrix and fitting linear model
```{r}
library(limma)

groups <- c("CON", "IR", "T2D")

subset <- plot_data %>% dplyr::select(geo_accession, group)
vec <- subset$group
big <- anno_expr %>% dplyr::select(one_of(subset$geo_accession))


f <- factor(as.character(vec), levels = groups)
design <- model.matrix(~0+f)
colnames(design) <- groups

fit <- lmFit(big,design)

con <- 
  makeContrasts(
    T2D-CON, # coef 1
    IR-CON, # coef 2
    levels = design
  )

fit <- contrasts.fit(fit, contrasts=con)
fit <- eBayes(fit)

limma::plotMA(fit, coef = 1) 
limma::plotMA(fit, coef = 2) 


```



## e) topTables
```{r}
library(limma)

gene_names <- anno %>% dplyr::select(ID = ENTREZID, gene_symbol = SYMBOL)

## TopTable ##
tt <- function(coef){
  topTable(fit,sort="p",n=Inf, coef = coef, resort.by = "p") %>% 
  rownames_to_column(var = "ID") %>% 
  left_join(gene_names, by = "ID") %>%
  dplyr::select(ID, gene_symbol, everything()) %>%
  filter(gene_symbol != "") %>%
  mutate(
    FC = 2^logFC,
    FC = ifelse(logFC<0,FC*(-1),FC),
    # lab_values = str_extract(gene_symbol, "MSTN|GDF8")
    )
}

topTable(fit,sort="p",n=Inf, coef = 1, resort.by = "p")

# CON v T2D, # coef 1
tt1 <- tt(1) 
# CON v IR, # coef 3
tt2 <- tt(2)


CONvT2D <- tt1 %>% filter(gene_symbol == "MSTN")
CONvIR <- tt2 %>% filter(gene_symbol == "MSTN")

tt_vanTienen <- 
  list(
  CONvT2D,
  CONvIR
  ) %>%  
  setNames(c("CONvT2D", "CONvIR")) %>%
    bind_rows(.id = "comparison")

```

## f) VolcanoPlot
```{r fig.height=6, fig.width=6}
library(EnhancedVolcano)

volc <- function(tt, title){
  EnhancedVolcano(tt,
    lab = tt$gene_symbol,
    x = 'logFC',
    y = 'P.Value',
    title = title,
    pCutoff = 10e-6,
    FCcutoff = 1,
    transcriptPointSize = 1.5,
    transcriptLabSize = 4.0,
    subtitle = "", 
    selectLab = tt$lab_values)
}

volc(tt1, "CON v T2D")
volc(tt2, "IR v T2D")

```

## g) Pull relevant data
```{r}

plot_data_vanTienen <- 
  plot_data %>% 
  dplyr::select(geo_accession, MSTN, group)

```

# 4. GSE25462 - Patti


## a) Extraction
```{r}
###########
## SETUP ##
###########

gse <- "GSE25462"

## Extraction of PROCESSED expressionSet ----
eset_proc <- getGEO(GEO = paste0(gse))
eset_proc <- eset_proc[[paste0(gse,"_series_matrix.txt.gz")]]

# Reading raw data
filePaths <-  getGEOSuppFiles(paste0(gse))
untar(paste0(gse,"/",gse,"_RAW.tar"), exdir=paste(gse,"data",sep = "/"))
cels <- list.files(paste0(gse,"/data/"), pattern = "[gz]")
sapply(paste(gse,"data", cels, sep="/"), gunzip)
celpath <- paste(gse,"data/", sep = "/")
fns <- list.celfiles(path=celpath,full.names=TRUE)
fns
cat("Reading files:\n",paste(fns,collapse="\n"),"\n")
 
#Loading the CEL-files into an AffyBatch object
celfiles <- ReadAffy(celfile.path=celpath)
oligo::boxplot(celfiles)

## RMA-normalization
eset <- rma(celfiles)
oligo::boxplot(eset)

## Extract expression matrix 
expr <- exprs(eset)
expr <- expr %>% as.data.frame %>% rownames_to_column(var = 'ID') # copy rownames to their own column

## Get annotation data for platform
anno_db <- select(hgu133plus2.db, keys = keys(hgu133plus2.db), columns=c("SYMBOL","GENENAME", "ENTREZID"), keytype="PROBEID") %>% 
  dplyr::rename(ID = PROBEID)


## Annotate data expressionSet
anno <-
  as.data.frame(eset@featureData@data) %>%
    rownames_to_column("ID") %>% 
    left_join(anno_db, by = "ID")
  
anno_expr <- dplyr::full_join(anno,expr, by = 'ID') # merge expr and anno
   
```

## b) Cleaning
```{r}

pData(eset) <- pData(eset_proc)


## EXTRACT/CLEAN PHENO DATA ##
pheno_eset <-
  pData(eset) %>% 
  dplyr::select(title) %>% 
  mutate(group = str_extract(title, "FamilyHistoryNegative|Muscle-Diabetes")) %>% 
  rownames_to_column("geo_accession") %>% 
  filter(!is.na(group)) %>% 
  mutate(group = recode(group, "Muscle-Diabetes" = "T2D", "FamilyHistoryNegative" = "CON"),
         geo_accession = paste0(geo_accession,".CEL")) 
  
  
pheno_eset
```

## c) Plot
```{r, fig.width = 10, fig.height=10}
## Set-up comparisons

groups <- pheno_eset %>% distinct(group) %>% pull

my_comparisons <- 
  list(
    c("CON", "T2D")
  )



## Plot data

plot_data <-
  anno_expr %>% 
  filter(str_detect(SYMBOL, "MSTN")) %>% 
  dplyr::select(GSM624925.CEL:GSM624974.CEL) %>% 
  t() %>% 
  as.data.frame %>% 
  rownames_to_column("geo_accession") %>% 
  dplyr::rename(MSTN = V1) %>% 
  left_join(pheno_eset, by = "geo_accession") %>% 
  filter(!is.na(group))

## PLOT ##
plot_data %>% 
  ggplot(aes(x = group, y = MSTN))+
   stat_summary(fun.data = mean_se, geom = "errorbar", width=0.2, size = 1, color = "#000000") +
   stat_summary(fun.y = mean, geom = "bar", width=0.8, size = 1, color = "#000000", fill=c("#ffffff", "black")) +
   theme_classic() +
   scale_y_continuous(expand = c(0, 0)) +
   scale_color_manual(values=c("#000000","#000000")) +
   ggbeeswarm::geom_quasirandom(size = 3, stroke = 1, width = 0.2, shape = 21, fill = "white")+
    theme(aspect.ratio = 1,
          plot.title = element_text(vjust= 7.5),
          axis.title.x = element_blank(),
          axis.text.x = element_text(color = "black"),
          axis.title.y = element_text(vjust = 4.5),
          axis.text.y = element_text(color = "black"),
          legend.position = "none",
          axis.ticks.x = element_blank(),
          plot.margin = unit(c(20,20,20,20),"pt")) + 
   labs(y=expression(paste(italic("MSTN "), log[2], " mRNA expression")),
        x="Group") + 
    ggpubr::stat_compare_means(aes(label = ..p.adj..), comparisons = my_comparisons, method = "t.test", hide.ns = T, p.adjust.method = "fdr") +
    coord_cartesian(clip = "off")

```

## d) Designing model matrix and fitting linear model
```{r}
library(limma)

groups <- c("CON", "T2D")

subset <- plot_data %>% dplyr::select(geo_accession, group)
vec <- subset$group
big <- anno_expr %>% dplyr::select(one_of(subset$geo_accession))


f <- factor(as.character(vec), levels = groups)
design <- model.matrix(~0+f)
colnames(design) <- groups

fit <- lmFit(big,design)

con <- 
  makeContrasts(
    T2D-CON, # coef 1
    levels = design
  )

fit <- contrasts.fit(fit, contrasts=con)
fit <- eBayes(fit)

limma::plotMA(fit, coef = 1) 


```



## e) topTables
```{r}
library(limma)

gene_names <- anno %>% dplyr::select(ID = ENTREZID, gene_symbol = SYMBOL)

## TopTable ##
tt <- function(coef){
  topTable(fit,sort="p",n=Inf, coef = coef, resort.by = "p") %>% 
  rownames_to_column(var = "ID") %>% 
  left_join(gene_names, by = "ID") %>%
  dplyr::select(ID, gene_symbol, everything()) %>%
  filter(gene_symbol != "") %>%
  mutate(
    FC = 2^logFC,
    FC = ifelse(logFC<0,FC*(-1),FC),
    # lab_values = str_extract(gene_symbol, "MSTN|GDF8")
    )
}


# CON v T2D, # coef 1
tt1 <- tt(1) 

CONvT2D <- tt1 %>% filter(gene_symbol == "MSTN")

tt1
CONvT2D

tt_Patti <- 
  list(
  CONvT2D
  ) %>%  
  setNames(c("CONvT2D")) %>%
    bind_rows(.id = "comparison")

```

## f) VolcanoPlot
```{r fig.height=6, fig.width=6}
library(EnhancedVolcano)

volc <- function(tt, title){
  EnhancedVolcano(tt,
    lab = tt$gene_symbol,
    x = 'logFC',
    y = 'P.Value',
    title = title,
    pCutoff = 10e-6,
    FCcutoff = 1,
    transcriptPointSize = 1.5,
    transcriptLabSize = 4.0,
    subtitle = "", 
    selectLab = tt$lab_values)
}

volc(tt1, "CON v T2D")

```

## g) Pull relevant data
```{r}

plot_data_Patti <- 
  plot_data %>% 
  dplyr::select(geo_accession, MSTN, group)

```

# 5. GSE121 - Yang

## a) Extraction
```{r}
###########
## SETUP ##
###########

## NOTE: This experiment is in a different format consisting of several 'subarrays'
## GPL100 series contains MSTN

gse <- "GSE121"

## Extraction of expressionSet and gene annotations ----
eset <- getGEO(GEO = paste0(gse))
eset <- eset[["GSE121-GPL100_series_matrix.txt.gz"]]

## QC ##
expr <- exprs(eset)
expr <- backgroundCorrect(expr, method= "normexp") # background correct
expr <- normalizeBetweenArrays(object = expr) # normalize expression profiles
expr <- log2(expr)
expr <- expr %>% as.data.frame %>% rownames_to_column(var = 'ID') 


# Boxplots
expr %>%
  gather(gsm, value, -ID) %>% 
  ggplot(aes(x = gsm, y = value))+
  geom_boxplot()
# Histogram
expr %>%
  gather(gsm, value, -ID) %>% 
  ggplot(aes(x =value))+
  geom_histogram()

## Annotate data expressionSet
anno <- 
  as.data.frame(eset@featureData@data) %>%  # extract annotation data
    dplyr::select(ID, `Gene Symbol`, ENTREZ_GENE_ID)

anno_expr <- dplyr::full_join(anno,expr, by = 'ID') # merge expr and anno

anno_expr %>% dplyr::filter(`Gene Symbol` == "MSTN")
```

## b) Cleaning
```{r}

## EXTRACT/CLEAN PHENO DATA ##
pheno_eset <-
  pData(eset) %>% 
  dplyr::select(title) %>% 
  mutate(group = str_extract(title, "insulin resistant|insulin sensitive")) %>% 
  rownames_to_column("geo_accession") %>% 
  mutate(group = recode(group, "insulin resistant" = "IR", "insulin sensitive" = "CON")) %>% 
  dplyr::select(-title)
  
  
pheno_eset
```

## c) Plot
```{r, fig.width = 10, fig.height=10}
## Set-up comparisons

groups <- pheno_eset %>% distinct(group) %>% pull

my_comparisons <- 
  list(
    c("CON", "IR")
  )



## Plot data

plot_data <-
  anno_expr %>% 
  filter(str_detect(`Gene Symbol`, "MSTN")) %>% 
  dplyr::select(GSM2287:GSM2331) %>% 
  t() %>% 
  as.data.frame %>% 
  rownames_to_column("geo_accession") %>% 
  dplyr::rename(MSTN = V1) %>% 
  left_join(pheno_eset, by = "geo_accession") %>% 
  filter(!is.na(group))

## PLOT ##

plot_data %>% 
  ggplot(aes(x = group, y = MSTN))+
   stat_summary(fun.data = mean_se, geom = "errorbar", width=0.2, size = 1, color = "#000000") +
   stat_summary(fun.y = mean, geom = "bar", width=0.8, size = 1, color = "#000000", fill=c("#ffffff", "grey60")) +
   theme_classic() +
   scale_y_continuous(expand = c(0, 0)) +
   scale_color_manual(values=c("#000000","#000000")) +
   ggbeeswarm::geom_quasirandom(size = 3, stroke = 1, width = 0.2, shape = 21, fill = "white")+
    theme(aspect.ratio = 1,
          plot.title = element_text(vjust= 7.5),
          axis.title.x = element_blank(),
          axis.text.x = element_text(color = "black"),
          axis.title.y = element_text(vjust = 4.5),
          axis.text.y = element_text(color = "black"),
          legend.position = "none",
          axis.ticks.x = element_blank(),
          plot.margin = unit(c(20,20,20,20),"pt")) + 
   labs(y=expression(paste(italic("MSTN "), log[2], " mRNA expression")),
        x="Group") + 
    ggpubr::stat_compare_means(aes(label = ..p.adj..), comparisons = my_comparisons, method = "wilcox.test", hide.ns = T, p.adjust.method = "fdr") +
    coord_cartesian(clip = "off")



```

## d) Designing model matrix and fitting linear model
```{r}
library(limma)

groups <- c("CON", "IR")

subset <- plot_data %>% dplyr::select(geo_accession, group)
vec <- subset$group
big <- anno_expr %>% dplyr::select(one_of(subset$geo_accession))

anno_expr %>% filter(`Gene Symbol` == "MSTN")

f <- factor(as.character(vec), levels = groups)
design <- model.matrix(~0+f)
colnames(design) <- groups

fit <- lmFit(big,design)

con <- 
  makeContrasts(
    IR-CON, # coef 1
    levels = design
  )

fit <- contrasts.fit(fit, contrasts=con)
fit <- eBayes(fit)

limma::plotMA(fit, coef = 1) 


```



## e) topTables
```{r}
library(limma)

gene_names <- anno %>% dplyr::select(ID = ENTREZ_GENE_ID, gene_symbol = `Gene Symbol`)

## TopTable ##
tt <- function(coef){
  topTable(fit,sort="p",n=Inf, coef = coef, resort.by = "p") %>% 
  rownames_to_column(var = "ID") %>% 
  # left_join(gene_names, by = "ID") %>%
  # dplyr::select(ID, gene_symbol, everything()) %>%
  # filter(gene_symbol != "") %>%
  mutate(
    FC = 2^logFC,
    FC = ifelse(logFC<0,FC*(-1),FC),
    # lab_values = str_extract(gene_symbol, "MSTN|GDF8")
    )
}




# CON v IR, # coef 1
tt1 <- tt(1) 


CONvIR <- tt1 %>% filter(ID == "2660")

tt1
CONvIR

tt_Yang <- 
  list(
  CONvIR
  ) %>%  
  setNames(c("CONvIR")) %>%
    bind_rows(.id = "comparison")

tt_Yang

```

## f) VolcanoPlot
```{r fig.height=6, fig.width=6}
library(EnhancedVolcano)

volc <- function(tt, title){
  EnhancedVolcano(tt,
    lab = tt$gene_symbol,
    x = 'logFC',
    y = 'P.Value',
    title = title,
    pCutoff = 10e-6,
    FCcutoff = 1,
    transcriptPointSize = 1.5,
    transcriptLabSize = 4.0,
    subtitle = "", 
    selectLab = tt$lab_values)
}

volc(tt1, "CON v IR")

```

## g) Pull relevant data
```{r}

plot_data_Yang <- 
  plot_data %>% 
  dplyr::select(geo_accession, MSTN, group)

```

# COMBINING DATA 
```{r}

## Create list of log expression values ----

data_list <- 
  list(
    plot_data_Gallagher,
    plot_data_Patti, 
    plot_data_vanTienen,
    plot_data_Wu,
    plot_data_Yang
  )

norm_list <- lapply(data_list, function(dat){
  
  con_mean <- 
    dat %>% 
    filter(group == "CON") %>% 
    summarise(mean = mean(MSTN, na.rm = T)) %>% pull

  norm_df <- 
    dat %>% 
    mutate(mean = con_mean,
           norm = MSTN-mean,
           log2_fc = norm/mean)
  
  return(norm_df)

}) %>% setNames(c("Gallagher", "Patti", "vanTienen", "Wu", "Yang"))


comb_df <- norm_list %>% bind_rows(.id = "id")


## PLOT ##

comb_df %>% 
  ggplot(aes(x = reorder(geo_accession, norm), y = norm)) +
  geom_segment(aes(x = reorder(geo_accession, norm), xend = reorder(geo_accession, norm), y = 0, yend = norm), color = "grey") +
  geom_point(aes(color = group), size=2) +
  theme_light() +
  theme(
    panel.grid.major.x = element_blank(),
    panel.border = element_blank(),
    axis.ticks.x = element_blank(),
    axis.text.x = element_blank()
  )

my_comparisons <- 
  list(
    c("CON", "IR"),
    c("IR", "T2D"),
    c("CON", "T2D")
  )

comb_df %>% 
  ggplot(aes(x = group, y = norm)) +
  ggbeeswarm::geom_quasirandom(aes(color = group), size=2) +
  stat_summary(fun.data = mean_se, geom = "errorbar", width = 0.1)+
  theme_light() +
  theme(
    panel.grid.major.x = element_blank(),
    panel.border = element_blank(),
    axis.ticks.x = element_blank()
  )+
  ggpubr::stat_compare_means(comparisons = my_comparisons, method = "wilcox.test")


## Combining log2FC values ----
comb_tt <- 
  list(
    tt_Gallagher,
    tt_Patti,
    tt_vanTienen,
    tt_Wu,
    tt_Yang
  ) %>% 
  setNames(c("Gallagher", "Patti", "vanTienen", "Wu", "Yang")) %>% 
  bind_rows(.id = "id")


final_plot_data <- 
  comb_tt %>% 
  mutate(id = factor(id, levels = c("Wu","vanTienen","Gallagher", "Patti", "Yang" )),
         comparison = recode(comparison, "CONvIR" = "Insulin-resistant", "CONvT2D" = "T2D"),
         comparison = factor(comparison, levels = c("Insulin-resistant", "T2D"))) 

final_plot_data %>%
  ggplot(aes(x = reorder(comparison, logFC), y = logFC))+
  geom_col(aes(fill = comparison), color = "black")+
  scale_y_continuous(limits = c(-1,1))+
  scale_fill_manual(name = "", values = c("white", "black"))+
  geom_text(aes(label = paste0("p=",round(adj.P.Val,2)),
               vjust = ifelse(logFC >= 0, -0.5, 1.5)),
            size = 3)+
  geom_hline(aes(yintercept = 0))+
  theme_classic()+
  labs(y=expression(paste(italic("MSTN "), log[2], " fold-change"))) + 
  theme(
    axis.text.x = element_blank(),
    axis.title.x = element_blank(),
    axis.ticks.x = element_blank(),
    axis.line.x = element_blank(),
    strip.background = element_blank(),
    panel.grid.major.y = element_line(color = "grey80"),
    panel.grid.minor.y = element_line(color = "grey90"),
    panel.spacing = unit(0, "lines")
  )+
  facet_grid(~id)


```

```{r}

```

